---
title: "non_censored_analysis"
author: "Naitong Chen, Shirley Cui, Shannon Edie"
date: "4/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r read.data.in, results='hide', warning=FALSE}
chardonnay2019 = read.csv("TanjaVoegel_Chardonnay_2019_data.csv")
chardonnay2020 = read.csv("TanjaVoegel_Chardonnay_2020_data.csv")
merlot2019 = read.csv("TanjaVoegel_Merlot_2019_data.csv")
merlot2020 = read.csv("TanjaVoegel_Merlot_2020_data.csv")
addition2021 = read.csv("data addition March 8 2021.csv")

# removing extra columns
merlot2020 = merlot2020[-(61:65),]
merlot2020 = merlot2020[,-(19:20)]
merlot2019 = merlot2019[-(61:65),]
merlot2019 = merlot2019[,-18]

# Make sure the merlot column names are consistent
colnames(merlot2020)[colnames(merlot2020)=="spad"] = "SPAD"

# filling in zero for yield related entries
merlot2020$yield[is.na(merlot2020$yield)] = 0
merlot2020$cluster.weight[is.na(merlot2020$cluster.weight)] = 0
merlot2020$berries.cluster[is.na(merlot2020$berries.cluster)] = 0
merlot2020$berry.weight[is.na(merlot2020$berry.weight)] = 0
merlot2020$cluster.number[is.na(merlot2020$cluster.number)] = 0

# encode row, treatment, block, and subsample as factor
chardonnay2019$row = as.factor(chardonnay2019$row)
chardonnay2019$treatment = as.factor(chardonnay2019$treatment)
chardonnay2019$block = as.factor(chardonnay2019$block)
chardonnay2019$subsample = as.factor(chardonnay2019$subsample)

chardonnay2020$row = as.factor(chardonnay2020$row)
chardonnay2020$treatment = as.factor(chardonnay2020$treatment)
chardonnay2020$block = as.factor(chardonnay2020$block)
chardonnay2020$subsample = as.factor(chardonnay2020$subsample)

merlot2019$row = as.factor(merlot2019$row)
merlot2019$treatment = as.factor(merlot2019$treatment)
merlot2019$block = as.factor(merlot2019$block)
merlot2019$subsample = as.factor(merlot2019$subsample)


merlot2020$row = as.factor(merlot2020$row)
merlot2020$treatment = as.factor(merlot2020$treatment)
merlot2020$block = as.factor(merlot2020$block)
merlot2020$subsample = as.factor(merlot2020$subsample)

library(plyr)
# Add in the additional data
names(merlot2020)[1] <- "Key"
merlot2020 <- join(merlot2020, addition2021)

# detach plyr as it creates conflict with tidyverse
detach(package:plyr)
# remove unused dataset
remove(addition2021)
```

```{r load.packages, echo=F, results='hide'}
library(tidyverse)
library(ggpubr) 
library(corrplot)
library(icenReg) # for survival analysis
library(knitr)
library(lawstat) # for the levene test
library(mice)
library(miceadds)
library(MKinfer)
library(rstatix) # for the repeated-measures ANOVA
library(ARTool) # for aligned rank transform
library(maditr) # for dcast
library(pwr2)
library(kableExtra)
library(magick)
library(car)
detach(package:car)
```

With the data loaded, we first conduct complete case analysis (ignoring missing data). We begin by checking the equal variance assumptions. The normality assumption will be checked once we get the residuals from the fitted ANOVA models. Note that the Levene test is less sensitive to violations of the normality assumption.

```{r chardonnay2019.variance.check}
resp.char19 <- colnames(chardonnay2019)[-c(1:5)]
homosc.table.char19 <- array(data=NA, dim=c(length(resp.char19), 2))
dimnames(homosc.table.char19)[[2]] <- c("Bartlett", "Levene")
dimnames(homosc.table.char19)[[1]] <- resp.char19

for (i in 1:length(resp.char19)) {
  # the lapply function simply extracts the p-value from each outputted run of the MICE data
  bart.p <- with(chardonnay2019, 
                               bartlett.test(eval(parse(text=resp.char19[i])) ~ paste(treatment, block)))$p.value
  lev.p  <- with(chardonnay2019, 
                        levene.test(eval(parse(text=resp.char19[i])), paste(treatment, block)))$p.value
  homosc.table.char19[i,1] <- bart.p
  homosc.table.char19[i,2] <- lev.p
}

for (i in 1:length(resp.char19)) {
  boxplot(chardonnay2019[[resp.char19[i]]] ~ chardonnay2019$treatment + chardonnay2019$block,
        xlab="Treatment + Block")
}

homosc.table.char19
```

```{r chardonnay2020.variance.check}
resp.char20 <- colnames(chardonnay2020)[-c(1:5)]
homosc.table.char20 <- array(data=NA, dim=c(length(resp.char20), 2))
dimnames(homosc.table.char20)[[2]] <- c("Bartlett", "Levene")
dimnames(homosc.table.char20)[[1]] <- resp.char19

for (i in 1:length(resp.char20)) {

  # the lapply function simply extracts the p-value from each outputted run of the MICE data
  bart.p <- with(chardonnay2020, 
                               bartlett.test(eval(parse(text=resp.char20[i])) ~ paste(treatment, block)))$p.value
  lev.p  <- with(chardonnay2020, 
                        levene.test(eval(parse(text=resp.char20[i])), paste(treatment, block)))$p.value
  homosc.table.char20[i,1] <- bart.p
  homosc.table.char20[i,2] <- lev.p
}

for (i in 1:length(resp.char20)) {
  boxplot(chardonnay2020[[resp.char20[i]]] ~ chardonnay2020$treatment + chardonnay2020$block,
        xlab="Treatment + Block")
}

homosc.table.char20
```

```{r merlot2019.variance.check}
resp.mer19 <- colnames(merlot2019)[-c(1:5, 7)]
# Create some tables to save the results of our statistical tests in
homosc.table.mer19 <- array(data=NA, dim=c(length(resp.mer19), 2))
dimnames(homosc.table.mer19)[[2]] <- c("Bartlett", "Levene")
dimnames(homosc.table.mer19)[[1]] <- resp.mer19

# Compare variances for each unique treatment-block group
for (i in 1:length(resp.mer19)) {

  # the lapply function simply extracts the p-value from each outputted run of the MICE data
  bart.p <- (with(merlot2019,
                        bartlett.test(eval(parse(text=resp.mer19[i])) ~ paste(treatment, block)))$p.value)
  lev.p  <- (with(merlot2019,
                        levene.test(eval(parse(text=resp.mer19[i])), paste(treatment, block)))$p.value)
  homosc.table.mer19[i,1] <- bart.p
  homosc.table.mer19[i,2] <- lev.p
}

for (i in 1:length(resp.mer19)) {
  boxplot(merlot2019[[resp.mer19[i]]] ~ merlot2019$treatment + merlot2019$block,
        xlab="Treatment + Block")
}

homosc.table.mer19
```

```{r merlot2020.variance.check}
resp.mer20 <- colnames(merlot2020)[-c(1:5, 7,9)]
# Create some tables to save the results of our statistical tests in
homosc.table.mer20 <- array(data=NA, dim=c(length(resp.mer20), 2))
dimnames(homosc.table.mer20)[[2]] <- c("Bartlett", "Levene")
dimnames(homosc.table.mer20)[[1]] <- resp.mer20

# Compare variances for each unique treatment-block group
for (i in 1:length(resp.mer20)) {

  # the lapply function simply extracts the p-value from each outputted run of the MICE data
  bart.p <- (with(merlot2020,
                        bartlett.test(eval(parse(text=resp.mer20[i])) ~ paste(treatment, block)))$p.value)
  lev.p  <- (with(merlot2020,
                        levene.test(eval(parse(text=resp.mer20[i])), paste(treatment, block)))$p.value)
  homosc.table.mer20[i,1] <- bart.p
  homosc.table.mer20[i,2] <- lev.p
}

for (i in 1:length(resp.mer20)) {
  boxplot(merlot2020[[resp.mer20[i]]] ~ merlot2020$treatment + merlot2020$block,
        xlab="Treatment + Block")
}

homosc.table.mer20
```

From above we see that the Levene tests do not show much major deviations from the equal variance assumptions. Bartlett tests results may not be so reliable here as with so few observations, we can easily violate the normality assumption. However, they serve as a good reference for comparions nonetheless. We now fit the two-way ANOVA models for response. As the same time, we show the distribution of the residuals and construct the 95% C.I. for the difference in treatment effects.

```{r two.way.anova.chardonnay2019}
resps = colnames(chardonnay2019)[6:8]

for (i in 1:length(resps)) {
  print(resps[i])
  options(contrasts = c("contr.sum", "contr.poly"))
  resres = aov(eval(parse(text=resps[i])) ~ treatment * block, data=chardonnay2019)

  plot(resres$residuals)
  qqnorm(resres$residuals)
  qqline(resres$residuals)
  
  options(contrasts = c("contr.sum", "contr.poly"))
  library(car)
  resresres = Anova(resres, type = "III")
  print(resresres)
  
  print(tukey_hsd(resres, which="treatment"))
}
```

```{r two.way.anova.chardonnay2020}
resps = colnames(chardonnay2020)[6:8]

for (i in 1:length(resps)) {
  print(resps[i])
  options(contrasts = c("contr.sum", "contr.poly"))
  resres = aov(eval(parse(text=resps[i])) ~ treatment * block, data=chardonnay2020)

  plot(resres$residuals)
  qqnorm(resres$residuals)
  qqline(resres$residuals)
  
  options(contrasts = c("contr.sum", "contr.poly"))
  library(car)
  resresres = Anova(resres, type = "III")
  print(resresres)
  
  print(tukey_hsd(resres, which="treatment"))
}
```

```{r two.way.anova.merlot2019}
resps = resp.mer19

for (i in 1:length(resps)) {
  print(resps[i])
  options(contrasts = c("contr.sum", "contr.poly"))
  resres = aov(eval(parse(text=resps[i])) ~ treatment * block, data=merlot2019)
  
  plot(resres$residuals)
  qqnorm(resres$residuals)
  qqline(resres$residuals)

  options(contrasts = c("contr.sum", "contr.poly"))
  library(car)
  resresres = Anova(resres, type = "III")
  print(resresres)
  
  print(tukey_hsd(resres, which="treatment")) 
}
```

```{r two.way.anova.merlot2020}
resps = names(merlot2020)[c(6,8,c(10:18))]

for (i in 1:length(resps)) {
  print(resps[i])
  options(contrasts = c("contr.sum", "contr.poly"))
  resres = aov(eval(parse(text=resps[i])) ~ treatment * block, data=merlot2020)
  
  plot(resres$residuals)
  qqnorm(resres$residuals)
  qqline(resres$residuals)
  
  options(contrasts = c("contr.sum", "contr.poly"))
  library(car)
  resresres = Anova(resres, type = "III")
  print(resresres)
  
  print(tukey_hsd(resres, which="treatment")) 
}
```

```{r two.way.merlot2020.pruning.weight.ravaz.index}
block12_removed = merlot2020[which(merlot2020$block != "1" & merlot2020$block != "2"),]

print("pruning.weight")
options(contrasts = c("contr.sum", "contr.poly"))
resres = aov(pruning.weight ~ treatment * block, data=block12_removed)

plot(resres$residuals)
qqnorm(resres$residuals)
qqline(resres$residuals)

options(contrasts = c("contr.sum", "contr.poly"))
library(car)
resresres = Anova(resres, type = "III")
print(resresres)

print(tukey_hsd(resres, which="treatment"))

print("Ravaz.index")
options(contrasts = c("contr.sum", "contr.poly"))
resres = aov(Ravaz.index ~ treatment * block, data=block12_removed)

plot(resres$residuals)
qqnorm(resres$residuals)
qqline(resres$residuals)

options(contrasts = c("contr.sum", "contr.poly"))
library(car)
resresres = Anova(resres, type = "III")
print(resresres)

print(tukey_hsd(resres, which="treatment"))
```

From above we see some of the normality assumptions are clearly violated. Again because we don't have many observations, it is difficult to tell whether the outliers causing these violations are part of the natual variation of the population. Therefore, instead of removing them, we run an alternative non-parametric model. Note that this only works for data without missing values. We will deal with those after the following non-paremetric tests.

```{r get.response.names}
resp.char19 <- colnames(chardonnay2019)[-c(1:5)]
resp.char20 <- colnames(chardonnay2020)[-c(1:5)]
# excluding veraison
resp.mer19 <- colnames(merlot2019)[-c(1:5, 7)]
# remove veraison and bloom
resp.mer20 <- colnames(merlot2020)[-c(1:5, 7,9)]
# responses with missing values
resp.mer20.miss <- resp.mer20[c(8:13)]
```

```{r art.anova.chardonnay2019}
resps = colnames(chardonnay2019)[6:8]
for (i in 1:length(resps)) {
  print(resps[i])
  
  model = art(eval(parse(text=resps[i])) ~ treatment * block, chardonnay2019)
  res = anova(model, type = 3)
  print(res)
}
```

```{r art.anova.chardonnay2020}
resps = colnames(chardonnay2020)[6:8]
for (i in 1:length(resps)) {
  print(resps[i])
  
  model = art(eval(parse(text=resps[i])) ~ treatment * block, chardonnay2020)
  res = anova(model, type = 3)
  print(res)
}
```

```{r art.anova.merlot2019}
resps = resp.mer19
for (i in 1:length(resps)) {
  print(resps[i])
  
  model = art(eval(parse(text=resps[i])) ~ treatment * block, merlot2019)
  res = anova(model, type = 3)
  print(res)
}
```

```{r art.anova.merlot2020}
resps = resp.mer20[1:7]
for (i in 1:length(resps)) {
  print(resps[i])
  
  model = art(eval(parse(text=resps[i])) ~ treatment * block, merlot2020)
  res = anova(model, type = 3)
  print(res)
}
```

These non-parametric results serve as a reference to check how sensitive the results are to the assumptions required in the two-way ANOVA models. We now address the missing data in the two merlot datasets. (Veraison and bloom as censored data will be treated in the censored analysis file)

```{r missing.rate.merlot2019}
table <- array(data=NA, dim=c(1,1))
rownames(table) <- c("veraison")
table[1,1] = paste0(round(mean(is.na(merlot2019$veraison)), 3)*100, "%")

print(table)
```

```{r missing.rate.merlot2020}
table <- array(data=NA, dim=c(1,7))
colnames(table) <- c("bloom", "veraison", "yield related measures", "berry quality related measures", 
                      "brown seed color", "pruning weight", "Ravaz index")
table[1,1] = paste0(round(mean(is.na(merlot2020$bloom)), 3)*100, "%")
table[1,2] = paste0(round(mean(is.na(merlot2020$veraison)), 3)*100, "%")
table[1,3] = paste0(round(mean(is.na(merlot2020$yield)), 3)*100, "%")
table[1,4] = paste0(round(mean(is.na(merlot2020$berry.Brix)), 3)*100, "%")
table[1,5] = paste0(round(mean(is.na(merlot2020$pruning.weight)), 3)*100, "%")
table[1,6] = paste0(round(mean(is.na(merlot2020$brown.seed.color)), 3)*100, "%")
table[1,7] = paste0(round(mean(is.na(merlot2020$Ravaz.index)), 3)*100, "%")

print(table)
```

We use multiple imputaion below to construct five imputed datasets for each response.

```{r multiple.imputation}
# filling in zero for yield related entries
merlot2020$yield[is.na(merlot2020$yield)] = 0
merlot2020$cluster.weight[is.na(merlot2020$cluster.weight)] = 0
merlot2020$berries.cluster[is.na(merlot2020$berries.cluster)] = 0
merlot2020$berry.weight[is.na(merlot2020$berry.weight)] = 0
merlot2020$cluster.number[is.na(merlot2020$cluster.number)] = 0

# imputation for merlot2020 using all responses (excluding key, row and subsample)
predMatrix = matrix(rep(1,ncol(merlot2020)^2), ncol(merlot2020), ncol(merlot2020))
diag(predMatrix) = 0
predMatrix[,1] = 0
predMatrix[,4] = 0
predMatrix[,5] = 0
merlot2020mice = mice(data = merlot2020, predictorMatrix = predMatrix, seed = 1, method = "norm.predict")
```

We now can run both the ANOVA and ART ANOVA models and pool the results together to obtain a new set of p-values. Again, these serve as references for checking how sensitive the complete case analyses are to the missing values.

```{r pooled.art.anova.merlot2020}
# make tables that will hold the p-values
resps = resp.mer20[8:13]
resp.mer20.miss = resp.mer20[8:13]
mer.2020.sig.miss.anova <- array(data=rep(0, length(resp.mer20.miss) * 3), dim=c(length(resp.mer20.miss), 3))
dimnames(mer.2020.sig.miss.anova)[[2]] <- c("treatment", "block", "interaction")
dimnames(mer.2020.sig.miss.anova)[[1]] <- resp.mer20.miss
mer.2020.sig.miss.art <- array(data=rep(0, length(resp.mer20.miss) * 3), dim=c(length(resp.mer20.miss), 3))
dimnames(mer.2020.sig.miss.art)[[2]] <- c("treatment", "block", "interaction")
dimnames(mer.2020.sig.miss.art)[[1]] <- resp.mer20.miss

for (i in 1:length(resps)) {
  print(resps[i])
  
  model_string = paste(resps[i], "~ treatment * block", sep=" ")
  res = mi.anova(merlot2020mice, model_string, type = 3)
  p.val = res$anova.table[["Pr(>F)"]][c(1:3)]
  mer.2020.sig.miss.anova[i,] = round(p.val,2)
  
  d1 = complete(merlot2020mice, 1)
  d2 = complete(merlot2020mice, 2)
  d3 = complete(merlot2020mice, 3)
  d4 = complete(merlot2020mice, 4)
  d5 = complete(merlot2020mice, 5)
  
  res1 = anova(art(eval(parse(text=resps[i])) ~ treatment * block, data = d1), type = 2)
  res2 = anova(art(eval(parse(text=resps[i])) ~ treatment * block, data = d2), type = 2)
  res3 = anova(art(eval(parse(text=resps[i])) ~ treatment * block, data = d3), type = 2)
  res4 = anova(art(eval(parse(text=resps[i])) ~ treatment * block, data = d4), type = 2)
  res5 = anova(art(eval(parse(text=resps[i])) ~ treatment * block, data = d5), type = 2)
  
  for (j in 1:3){
    fval = c(
      res1[["F value"]][j],
      res2[["F value"]][j],
      res3[["F value"]][j],
      res4[["F value"]][j],
      res5[["F value"]][j]
    )
    comb.p = micombine.F(fval, res1[["Df"]][j])
    mer.2020.sig.miss.art[i,j] = round(comb.p[["p"]],2)
  }
}

# results are summarized in these two tables (see R Console output)
print(mer.2020.sig.miss.anova)
print(mer.2020.sig.miss.art)
```

We now combine both years together by introducing year as a third factor.

```{r three.way.anova.chardonnay}
chardonnay2019$year <- 2019
chardonnay2020$year <- 2020
chardonnay2019$year <- as.factor(chardonnay2019$year)
chardonnay2020$year <- as.factor(chardonnay2020$year)
chardonnay.bothyr <- rbind(chardonnay2019, chardonnay2020)
names(chardonnay.bothyr)[[1]] <- "Key"
chardonnay.bothyr$Key <- as.factor(chardonnay.bothyr$Key)

for (i in c(6:8)) {
  print(names(chardonnay.bothyr)[i])
  res = aov(eval(parse(text=names(chardonnay.bothyr)[i])) ~ block*treatment*year,
            data=chardonnay.bothyr)
  resres = Anova(res, type = "III")
  print(resres)

  print(tukey_hsd(res, which="treatment"))
  
  res.art = anova(art(eval(parse(text=names(chardonnay.bothyr)[i])) ~ treatment * block * year, 
                  data = chardonnay.bothyr), type = 3)
  print(res.art)
}
```

```{r three.way.anova.merlot}
merlot2019$year <- 2019
merlot2020$year <- 2020
merlot2019$year <- as.factor(merlot2019$year)
merlot2020$year <- as.factor(merlot2020$year)
names(merlot2019)[[1]] <- "Key"
merlot.bothyr <- rbind(merlot2019[,c(1:6, 8:18)], merlot2020[,c(1:5,8, 10:17, 19:21)])
merlot.bothyr[117,1] <- "M57"
merlot.bothyr$Key <- as.factor(merlot.bothyr$Key)

for (i in c(6:16)) {
  print(names(merlot.bothyr)[i])
}

for (i in c(6:16)) {
  print(names(merlot.bothyr)[i])
  dat = merlot.bothyr[,c(1:3,i,17)]
  inds = which(is.na(merlot.bothyr[,i]))
  if (length(inds) > 0) {
    inds.m60 = inds - 60 
    dat = dat[-c(inds, inds.m60),]
    if (i == 15 | i == 16) {
      dat = dat[-which(dat$block == 2),]
    }
  }
  options(contrasts = c("contr.sum", "contr.poly"))
  res = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ block*treatment*year, data=dat)
  resres = Anova(res, type = "III")
  print(resres)
  
  print(tukey_hsd(res, which="treatment"))
}
```

Again as a comparison, we run the non-parametric version of the test (on responses without missing data). Missing data again is treated afterwards.

```{r three.way.art.chardonnay}
for (i in c(6:8)) {
  print(names(chardonnay.bothyr)[i])
  res.art = anova(art(eval(parse(text=names(chardonnay.bothyr)[i])) ~ treatment * block * year, 
                  data = chardonnay.bothyr), type = 3)
  print(res.art)
}
```

```{r three.way.art.merlot}
for (i in c(6:11)) {
  print(names(merlot.bothyr)[i])
  dat = merlot.bothyr[,c(1:3,i,17)]
  options(contrasts = c("contr.sum", "contr.poly"))
  res.art = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = merlot.bothyr), type = 3)
  print(res.art)
}
```

Finally, we address the missing data the same way as we did before and check both the results from parametric and non-parametric models.

```{r get.combined.datasets.ready}
merlot2019$year <- 2019
    complete_merlot2020_1 = complete(merlot2020mice, 1)
    complete_merlot2020_1$year <- 2020
    merlot2019$year <- as.factor(merlot2019$year)
    complete_merlot2020_1$year <- as.factor(complete_merlot2020_1$year)
    names(merlot2019)[[1]] <- "Key"
    merlot.bothyr.1 <- rbind(merlot2019[,c(1:6, 8:18)], complete_merlot2020_1[,c(1:5,8, 10:17, 19:21)])
    merlot.bothyr.1[117,1] <- "M57"
    merlot.bothyr.1$Key <- as.factor(merlot.bothyr.1$Key)
    
    merlot2019$year <- 2019
    complete_merlot2020_2 = complete(merlot2020mice, 2)
    complete_merlot2020_2$year <- 2020
    merlot2019$year <- as.factor(merlot2019$year)
    complete_merlot2020_2$year <- as.factor(complete_merlot2020_2$year)
    names(merlot2019)[[1]] <- "Key"
    merlot.bothyr.2 <- rbind(merlot2019[,c(1:6, 8:18)], complete_merlot2020_2[,c(1:5,8, 10:17, 19:21)])
    merlot.bothyr.2[117,1] <- "M57"
    merlot.bothyr.2$Key <- as.factor(merlot.bothyr.2$Key)
    
    merlot2019$year <- 2019
    complete_merlot2020_3 = complete(merlot2020mice, 3)
    complete_merlot2020_3$year <- 2020
    merlot2019$year <- as.factor(merlot2019$year)
    complete_merlot2020_3$year <- as.factor(complete_merlot2020_3$year)
    names(merlot2019)[[1]] <- "Key"
    merlot.bothyr.3 <- rbind(merlot2019[,c(1:6, 8:18)], complete_merlot2020_3[,c(1:5,8, 10:17, 19:21)])
    merlot.bothyr.3[117,1] <- "M57"
    merlot.bothyr.3$Key <- as.factor(merlot.bothyr.3$Key)
    
    merlot2019$year <- 2019
    complete_merlot2020_4 = complete(merlot2020mice, 4)
    complete_merlot2020_4$year <- 2020
    merlot2019$year <- as.factor(merlot2019$year)
    complete_merlot2020_4$year <- as.factor(complete_merlot2020_4$year)
    names(merlot2019)[[1]] <- "Key"
    merlot.bothyr.4 <- rbind(merlot2019[,c(1:6, 8:18)], complete_merlot2020_4[,c(1:5,8, 10:17, 19:21)])
    merlot.bothyr.4[117,1] <- "M57"
    merlot.bothyr.4$Key <- as.factor(merlot.bothyr.4$Key)
    
    merlot2019$year <- 2019
    complete_merlot2020_5 = complete(merlot2020mice, 5)
    complete_merlot2020_5$year <- 2020
    merlot2019$year <- as.factor(merlot2019$year)
    complete_merlot2020_5$year <- as.factor(complete_merlot2020_5$year)
    names(merlot2019)[[1]] <- "Key"
    merlot.bothyr.5 <- rbind(merlot2019[,c(1:6, 8:18)], complete_merlot2020_5[,c(1:5,8, 10:17, 19:21)])
    merlot.bothyr.5[117,1] <- "M57"
    merlot.bothyr.5$Key <- as.factor(merlot.bothyr.5$Key)
```

```{r table.for.three.way.merlot.missing}
# make tables that will hold the p-values
resp.mer20.miss = names(merlot.bothyr)[12:16]
mer.2020.sig.miss.anova <- array(data=rep(0, length(resp.mer20.miss) * 3), dim=c(length(resp.mer20.miss), 3))
dimnames(mer.2020.sig.miss.anova)[[2]] <- c("treatment", "block", "interaction")
dimnames(mer.2020.sig.miss.anova)[[1]] <- resp.mer20.miss
mer.2020.sig.miss.art <- array(data=rep(0, length(resp.mer20.miss) * 3), dim=c(length(resp.mer20.miss), 3))
dimnames(mer.2020.sig.miss.art)[[2]] <- c("treatment", "block", "interaction")
dimnames(mer.2020.sig.miss.art)[[1]] <- resp.mer20.miss
```

```{r pooled.three.way.merlot}
  for (i in c(12:16)) {
    print(names(merlot.bothyr)[i])
    dat.1 = merlot.bothyr.1[,c(1:3,i,17)]
    dat.2 = merlot.bothyr.2[,c(1:3,i,17)]
    dat.3 = merlot.bothyr.3[,c(1:3,i,17)]
    dat.4 = merlot.bothyr.4[,c(1:3,i,17)]
    dat.5 = merlot.bothyr.5[,c(1:3,i,17)]

    options(contrasts = c("contr.sum", "contr.poly"))
    res1 = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.1), type = 3)
    res2 = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.2), type = 3)
    res3 = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.3), type = 3)
    res4 = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.4), type = 3)
    res5 = anova(art(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.5), type = 3)

    for (j in c(1,2,4)){
      fval = c(
        res1[["F value"]][j],
        res2[["F value"]][j],
        res3[["F value"]][j],
        res4[["F value"]][j],
        res5[["F value"]][j]
      )
      comb.p = micombine.F(fval, res1[["Df"]][j])
      if (j == 4) {
        mer.2020.sig.miss.art[i-11,j-1] = round(comb.p[["p"]],2) 
      } else {
        mer.2020.sig.miss.art[i-11,j] = round(comb.p[["p"]],2) 
      }
    }
    
    options(contrasts = c("contr.sum", "contr.poly"))
    res1 = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.1)
    res2 = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.2)
    res3 = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.3)
    res4 = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.4)
    res5 = aov(eval(parse(text=names(merlot.bothyr)[i])) ~ treatment * block * year, 
                  data = dat.5)
               
    for (j in c(1,2,4)){
      fval = c(
        summary(res1)[[1]][["F value"]][j],
        summary(res2)[[1]][["F value"]][j],
        summary(res3)[[1]][["F value"]][j],
        summary(res4)[[1]][["F value"]][j],
        summary(res5)[[1]][["F value"]][j]
      )
      comb.p = micombine.F(fval, summary(res1)[[1]][["Df"]][j])
      if (j == 4) {
        mer.2020.sig.miss.anova[i-11,j-1] = round(comb.p[["p"]],2) 
      } else {
        mer.2020.sig.miss.anova[i-11,j] = round(comb.p[["p"]],2) 
      }
    }
  }

print(mer.2020.sig.miss.anova)
print(mer.2020.sig.miss.art)
```